How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-04-18 1.01-beforeyoustart.Rmd
Add files via upload
# some changes after reading Parts I and II

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you are going to do afterwards. There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make the informed decisions from the outset. For example, reconciling yourself to using the statistical package R will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on. Getting, and staying in, the right mental framework will also be very important for your ability to work under pressure. Again, getting into good habits at the beginning will put you in a commanding position later.

@@

@@

PhDs are granted by universities (for the most part) and those are the institutions that set the standards. In many countries, these universities are required to meet the criteria of national bodies. Essentially, the top-down control is to ensure that the PhD qualification adheres to national and international standards. You may be aware that there are mutterings of whether qualifications are what they used to be. This has also been said of PhDs. However, I'm of the opinion that the standards are much as they were. You will really set the level on the bar of your own PhD. You either pass or fail. If your intention were to do the 'bare minimum' required to get a PhD, then you would have to look carefully for an advisor that would be happy to preside over such a thesis. Suffice it to say that this would be a very poor place to start, and there may be lots of other reasons why doing a PhD really is not appropriate for you.

## Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

@@

I can tell you that many examiners will treat the exercise the same irrespective of the questions asked of them. They will treat the examination as an extra-large task in peer review. This is not unfair. Personally, 
I think that there are some special extra hurdles required for a PhD (over and above 5 chapters that are publishable):

@@

I prefer a thesis that fits together within a stated framework. I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their proposal. A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy]. In this approach, you are encouraged to consider the 'bigger question' in your research area, identifying both what studies have produced evidence and identifying gaps from a theory driven approach. Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD even as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

@@

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. if the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

@@

Depending on where you study might mean that you have different amounts of course time. There are some institutions that will carry on funding you for as long as you want to carry on studying. This is unusual. Most countries now have a restriction on the amount of time that you can spend studying a PhD.

It does make a difference where in the world you do your PhD. There are some beautiful places to work. Some places are more expensive to live and therefore survive. If you are a biologist then you might want to consider where your study animals or plants are in order to decide your place to study. But most of all your choice of where to study is likely to be dictated by the presence of an adviser that you want to work with, and who wants you as their PhD student. Advisors are tied to their institutions, and any principal advisor will need to register you at their own institution, meaning that this will be where you spend your desk time. Some advisors might be more flexible than others, but it will be most likely that you will need to move to the institution where your advisor is based.

@@

The advisor that you're looking for should be offering or doing projects in which you are interested. There may only be a few people in the world that are doing this type of research, and you may end up having to choose between them. Hence, while you're reading about how to pick a good advisor, your shortlist may only be of two or three people in the world. The project and subject area are going to be strong areas driving your choices moving forwards.

One of the best ways of picking a good advisor is by talking to current or past students. You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects. A good adviser should have an open door so that you can talk to them whenever you need to. Thus, once you have talked to the students in their group, you should also make an appointment to talk to them about your project. It might not be one meeting, there may be several. The more contact you have with them, the more likely you are to be sure that this is the person whose help that you want when conducting research over the next few years.

@@

A good adviser should have a good network. Both during your PhD studies and after them, you will rely primarily on the network of your advisor as a basis to building your own network. For example, you will go to meetings that they go to, you will be introduced to people that they know, and you may have opportunities to collaborate with people within their networks [see @adams2012rise]. Doing a PhD with an advisor that has no network and that does not link up with other academics (or worse is disliked by other academics) might mean that your own work is not recognised, or even sidelined. You can easily check on the network of a potential advisor by looking at the collaborators in their publications, or on their website.

Having said this, the best academics also tend to be very busy, and that may mean that they do not make good advisors. The balance is definitely very tricky, and one of the best ways to make sure that you have enough time for all of your students is to limit the number of students you have. Clearly, this works against prospective students.

## What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisers who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided.

@@

## What makes a good project?
After having made lots of statements about how a project should be your own project the following advice might seem a bit strange. Clearly the work that you're going to be doing will need to fit into the group of the adviser that you've chosen.

The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for at least 3 years.

>>>>

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards. There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset. For example, reconciling yourself to using the [statistical package R](useR.html) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on. Preparing yourself into the right mental framework will also be very important for your ability to work under pressure. Getting into good habits at the beginning will put you in a commanding position later. The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.

@@

Universities are worried about two major aspects of those who start PhD studies:

 - The median time to completion in many countries worldwide is much longer than it should be [e.g. 7.5 years in the USA @geven2018how].
 - Completion rates are shockingly low [as low as 50% in the USA @litalien2015improving].

This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies. But there is also the need for students to comprehend right from the start that they are undertaken a major commitment that will likely impact many different parts of their lives.

@@

PhDs are granted by universities (for the most part) and those are the institutions that set the standards. In many countries, these universities are required to meet the criteria of national bodies. Essentially, the top-down control is to ensure that the PhD qualification adheres to national and international standards. You may be aware that there are mutterings of whether qualifications are what they used to be. This has also been said of PhDs. However, I'm of the opinion that the standards are much as they were. You will really set the level on the bar of your own PhD. You either pass or fail. If your intention were to do the 'bare minimum' required to get a PhD, then you would have to look carefully for an advisor that would be happy to preside over such a thesis. Suffice it to say that this would be a very poor place to start, and if this were your attitude there may be lots of other reasons why doing a PhD really is not appropriate.

## Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area, and learn to communicate this to different types of audience. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

@@

I can tell you that many examiners will treat the exercise the same irrespective of the questions asked of them. They will treat the examination as an extra-large task in peer review. This is not unfair. Personally, I think that there are some special extra hurdles required for a PhD (over and above 5 data chapters that are publishable):

@@

I prefer a thesis that fits together within an explicit [conceptual framework](PhDintro.html). I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](proposal.html). A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy]. In this approach, you are encouraged to consider the '[bigger question](bigidea.html)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach. Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

@@

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work [see @chamberlain2016ten]. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. if the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

@@

Depending on where you study might mean that you have different amounts of course time. There are some institutions that will carry on funding you for as long as you want to carry on studying. This is unusual. Most countries now have a restriction on the amount of time that you can spend studying a PhD, and given that completion times are so long [@litalien2015improving], there will be added pressure to finish. It may be important to consider the institutional support given to postgraduate students, outside of their research sphere.

It does make a difference where in the world you do your PhD. There are some beautiful places to work. Some places are more expensive to live and therefore survive. If you are a biologist then you might want to consider where your study animals or plants are in order to decide your place to study. But most of all your choice of where to study is likely to be dictated by the presence of an advisor that you want to work with, and who wants you as their PhD student. Advisors are tied to their institutions, and any principal advisor will need to register you at their own institution, meaning that this will be where you spend your desk time. Some advisors might be more flexible than others, but it will be most likely that you will need to move to the institution where your advisor is based.

@@

The advisor that you're looking for should be offering or doing projects in which you are interested. There may only be a few people in the world that are doing this type of research, and you may end up having to choose between them. Hence, while you're reading about how to pick a good advisor, your shortlist may only be of two or three people in the world. The project and subject area are going to be strong reasons driving your choices moving forwards.

One of the best ways of picking a good advisor is by talking to current or past students. You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects. A good advisor should have an open door so that you can talk to them whenever you need to. Thus, once you have talked to the students in their group, you should also make an appointment to talk to them about your project. It might not be one meeting, there may be several. The more contact you have with them, the more likely you are to be sure that this is the person whose help and advice that you want when conducting research over the next few years.

@@

A good advisor should have a good network. Both during your PhD studies and after them, you will rely primarily on the network of your advisor as a basis to building your own network. For example, you will go to meetings that they go to, you will be introduced to people that they know, and you may have opportunities to collaborate with people within their networks [see @adams2012rise]. Doing a PhD with an advisor that has no network and that does not link up with other academics (or worse is disliked by other academics) might mean that your own work is not recognised, or even sidelined. You can easily check on the network of a potential advisor by looking at the collaborators in their publications, or on their website.

Having said this, the best academics also tend to be very busy, and that may mean that they do not make good advisors. The balance is definitely very tricky, and one of the best ways for advisors to make sure that you have enough time for all of your students is to limit the number of students you have. Clearly, this works against prospective students.

## What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisors who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided. Chamberlain [-@chamberlain2016ten] suggests that there are 10 types of academic advisors, implying that 9 of these are rather disfunctional.

@@

## What makes a good project?
After having made lots of statements about how a project should be your own project the following advice might seem a bit strange. Clearly the work that you're going to be doing will need to fit into the group of the advisor that you've chosen.

The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for ~3 years.

### 2021-04-18 1.02-useR.Rmd
Add files via upload
# some changes after reading Parts I and II

# Reconciling yourself to doing things you've been avoiding {#useR}
Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. Each time the cursor got to the end of the screen I hit the carriage return button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.

@@

If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal. The real point here is not that you can only use R, but that R represents a platform that can be easily reproduced (because it’s free unlike many statistical software packages), and by saving and making the code available, you are making your data analyses reproducible. We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and therefore best practice.

>>>>

# Reconciling yourself to doing things you've been avoiding {#useR}
Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the carriage return button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.

@@

If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal. The real point here is not that you can use R, but that R represents a platform on which analyses can be easily repeated, and by saving and making the code available, you are making your data analyses reproducible. We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and therefore best practice.

### 2021-04-18 2.01-Hyp.Rmd
Add files via upload
# some changes after reading Parts I and II

The writing part of the PhD is often the most difficult for many students. I think that there are several reasons for this. Writing well comes with practice, and most PhD students have had little experience with academic writing - even if they have already completed a MSc. The same could be said of most early career scientists (even after their PhDs), because writing takes practice and we shouldn't expect that it'll come automatically or immediately. In order to get this practice you need to get started, and (again) many students find starting to write challenging. Here is where this guide will help.

@@

Before you start writing in earnest, it’d be good to read this part of the book and make sure that you are familiar with the ideas therein. Later, you may need to refer back to sections once you’ve started writing.

>>>>

The writing part of the PhD is often the most difficult for many students. I think that there are several reasons for this. Writing well comes with practice, and most PhD students have had little experience with academic writing - even if they have already completed a MSc. The same could be said of most early career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months. YOu shouldn't expect that scientific writing will come automatically or immediately. In order to get this practice you need to get started, and (again) many students find starting to write challenging. This guide was written to help you get started.

@@

Before you start writing in earnest, it’d be good to read this part of the book and make sure that you are familiar with the ideas herein. Later, you may need to refer back to sections once you’ve started writing.

### 2021-04-18 2.08-citations.Rmd
Add files via upload
# some changes after reading Parts I and II

## Why do you need to cite? {#citations}

@@

### Research is built on existing work and ideas

@@

### Your citations say a lot about you

@@

### Cite while you write (not afterwards)

@@

### Citation Styles

@@

#### Vancouver Style

Vancouver style dates back to medical journals in the 1970s and refer to conventions to avoid fraud in medical sciences. In biological sciences, they are not used as widely as Harvard style, but are chosen as they don’t impede the flow of the text for the reader.

@@

#### Harvard Style

@@

### Moving from Harvard to Vancouver

Essentially, every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation. However, Vancouver style does not allow for much variation in how you cite. You can’t use (e.g. ^1,2^) or (see ^3^ for a review). This means that when you originally wrote for a Harvard style journal, but then change the manuscript to submit it to a journal that uses Vancouver style, you’ll need to remove any explanations of citations. That doesn’t mean removing the citations themselves, but it may mean re-writing some sentences. Simply changing your reference manager to use the different referencing style will only be the start of your work.

### Where within a sentence should the citation come?

@@

### What about the order of the citations in a string?

@@

### What about citations as taxonomic authorities?

@@

### Is it possible to mis-cite?

@@

### Should I cite without reading the paper?

@@

### What should I not cite?

@@

### Do I cite the review or the primary literature?

@@

### What is primacy and why does it matter?

@@

### How many citations is enough?

@@

### Should I cite myself?

@@

### Should I cite my friends?

@@

### Does the impact factor of the cited article matter?

>>>>

# Why do you need to cite? {#citations}

@@

## Research is built on existing work and ideas

@@

## Your citations say a lot about you

@@

## Cite while you write (not afterwards)

@@

## Citation Styles

@@

### Vancouver Style

Vancouver style dates back to medical journals in the 1970s and refer to conventions to avoid fraud in medical sciences. You are probably familiar with it even if you don't know the name. It is the convention that uses small superscript numbers to denote the place of citations in the main text, and the full reference given at the end of the document is provided in order of the citations (and their numbers), instead of alphabetical order. In biological sciences, Vancouver style is not used as widely as Harvard style, but is chosen as the superscript numbers tend not to impede the flow of the text for the reader.

@@

### Harvard Style

@@

## Moving from Harvard to Vancouver

Moving within styles is always a hassle, especially as formatting the references can take a long time, but moving between these styles is especially tricky as it may require you to re-evaluate the way in which you use the citations.

Moving from Harvard to Vancouver style means that every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation. However, Vancouver style does not allow for much variation in how you cite. You can’t use (e.g. ^1,2^) or (see ^3^ for a review). This means that when you originally wrote for a Harvard style journal, but then change the manuscript to submit it to a journal that uses Vancouver style, you’ll need to remove any explanations of citations. That doesn’t mean removing the citations themselves, but it may mean re-writing some sentences. Simply changing your reference manager to use the different referencing style will only be the start of your work.

## Where within a sentence should the citation come?

@@

## What about the order of the citations in a string?

@@

## What about citations as taxonomic authorities?

@@

## Is it possible to mis-cite?

@@

## Should I cite without reading the paper?

@@

## What should I not cite?

@@

## Do I cite the review or the primary literature?

@@

## What is primacy and why does it matter?

@@

## How many citations is enough?

@@

## Should I cite myself?

@@

## Should I cite my friends?

@@

## Does the impact factor of the cited article matter?

### 2021-04-18 2.09-databases.Rmd
Add files via upload
# some changes after reading Parts I and II

## Literature databases {#databases}

Databases of scientific literature have been around since the 1960s when the number of journals and papers started increasing so much that scholars simply couldn’t keep up. Quite simply, you should use one or more of these databases to look for relevant literature to read. The first database, and what continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of journals (titles, abstracts, authors and other metadata), but also the citations that each paper used. As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore a highly effective way of finding related and pertinent literature in your field.

@@

### Searching the literature using a database

There are two typical ways in which you can search for literature about your area of interest. The first is to use keywords and the second is to use citations of key references. I suggest that you use both of these tactics when doing your initial literature search. Further, I’d suggest that it’s worth your time keeping a record of what you have searched for, how and when and then download the results. This is called a systematic literature review. Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first introductory chapter for your thesis.

### Keywords

While searching for keywords sounds easy, picking the keywords that you should use is more difficult and requires a lot of practice with the databases and your key references. Remember that keywords are often not single words, but also include short phrases. Look at all of the key references and the keywords that they contain. Any keywords that they share are going to be important for you. Most literature databases have the option of using keywords or keywords and words in the title and abstract. I would start with the broader search term and see how well they work. However, these searches alone may return such a large amount of literature that you can’t possibly read it all. In this case, you will need to use more than a single keyword. The trick is to find the correct combination of keywords that will deliver all (or most) of your key references, while not delivering returns of thousands of papers that are not relevant to your study. Ideally, you want a search term that will deliver around one hundred papers at most, where the majority have relevance to your study area. Luckily, literature databases have some tricks for you to add to combine your keywords in ways that you can include or exclude certain topics.

#### Boolean terms or operators

@@

- Use AND when you want both terms included. For example, *performance AND lizards*. In this example, all papers that mention performance AND lizards in the title, abstract or keywords are returned. 
- Use OR when you want to include both terms. For example, *performance AND lizards OR geckos*. This example should return more literature than the previous search as it will include, all papers that mention *performance AND lizards OR geckos* in the title, abstract or keywords are returned. You will note that geckos are lizards, but people who write about geckos might not mention that they are lizards in the title, abstract or keywords, so you may need to use these extra search terms in order to pick up all of the relevant literature on lizards. If your research requires you to cover all lizards, then you might need a lot of search terms in the OR string in order to pick up all of the literature that covers lizard performance.
- Use NOT when you specifically want to exclude a keyword. Following our example above, the search term *performance AND reptiles NOT snakes* should give you all performance papers on lizards, but you may want to exclude turtles and crocodilians as well if they have a lot of hits.

#### Proximity operators and parentheses

You can also use the term NEAR to indicate that two terms need to occur within a specified number of words. For example, performance NEAR/5 reptile will capture items where these occur within 5 words of each other, but no greater. This might be useful to exclude some papers that mention performance and reptiles in the abstract, but are not about reptile performance.
Once you start putting together search strings you will need to use brackets to make sure that the correct terms are grouped. For example, performance AND (lizards OR geckos) will get what we described above, whereas (performance AND lizards) OR geckos will give you all papers on lizards and performance plus all papers on geckos (not what we wanted).

@@

#### Wildcards

@@

#### Combining searches

@@

#### Citations

@@

What else are literature databases used for?
The Science Citation Index used to be published as a single volume that emerged each year and was available in the reference section of the library. By the time that I started at university as an undergraduate in the 1980s, it was several volumes large and my first task for my tutor (one Dr Dave Thompson) was to look up all of his citations. Within a couple of years, the database was digitised and available on CD, and eventually became accessible through the internet.

@@

### Administrators want metrics

@@

### Google Scholar, Web of Science or Scopus?

@@

I have been known to take the odd peak at my Google Scholar profile over the year, and see how it's coming along. I rarely check on WoS or Scopus, 'cos it's a bit of a faff getting signed in and doing the search. Plus it looks so much smaller when one is habituated to seeing those double digits in GS! However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among others that they give) as I've never really known what all that extra represents. Something grey and unseemly? Well, it turns out that it's all good stuff, and perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.

It is worth noting that of the three indices, Google Scholar seems to be the easiest to manipulate. Delgado López‐Cózar et al [-@delgadolopez-cozar2014google] show how it is possible to manipulate citations and therefore metrics of individual researchers by depositing false documents into institutional repositories (note that they did this as an experiment, and you may find that restrictions on doing this at your own institution are prohibitive or lax). The authors also note that a consistent problem with Google Scholar is the lack of transparency, and therefore verification, of how the bibliometrics are calculated. This will make a difference to what papers are found, as Google Scholar tends to report those with the greatest citations first on the first page of outputs. Hence once a paper is initially boosted by manipulation, it will be more likely to be found and increasingly cited.

@@

### What’s the difference between indices?

Martín-Martín et al [-@martin-martin2018google] allows you to interpret your GS results as a more inclusive citation index. While WoS and Scopus aren't exclusively English or journal publications, they are mostly. But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream. Is your GS score more than a third higher than your WoS or Scopus score? If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.

The excerpts from the two tables above show how well GS correlates with both WoS and Scopus in our area (Biological Sciences). It also tells you by how much the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus. Again, if you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.

>>>>

# Literature databases {#databases}

Databases of scientific literature have been around since the 1960s when the number of journals and papers started increasing so much that scholars simply couldn’t keep up. Quite simply, you should use one or more of these databases to look for relevant literature to read. The first database, and what continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used. As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore a highly effective way of finding related and pertinent literature in your field.

@@

## Searching the literature using a database

There are two typical ways in which you can search for literature about your area of interest. The first is to use keywords and the second is to use citations of key references. I suggest that you use both of these tactics when doing your initial literature search. Further, I’d suggest that it’s worth your time keeping a record of what you have searched for, how and when and then download the results. This is called a systematic literature review. Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first [introductory chapter](PhDintro.html) for your thesis.

## Keywords

While searching for keywords sounds easy, picking the keywords that you should use is more difficult and requires a lot of practice with the databases and your key references. Remember that keywords are often not single words, but also include short phrases. Look at all of the key references and the keywords that they contain. Any keywords that they share are going to be important for you. Make a note of all relevant keywords for your searches, as ultimately this will help you when you need to determine the keywords for [your own chapters](title.html).

Most literature databases have the option of using just keywords, or keywords and words in the title and abstract. I would start with the broader search term and see how well they work. However, these searches alone may return such a large amount of literature that you can’t possibly read it all. In this case, you will need to use more than a single keyword. The trick is to find the correct combination of keywords that will deliver all (or most) of your key references, while not delivering returns of thousands of papers that are not relevant to your study. Ideally, you want a search term that will deliver around one hundred papers at most, where the majority have relevance to your study area. Luckily, literature databases have some tricks for you to add to combine your keywords in ways that you can include or exclude certain topics.

### Boolean terms or operators

@@

- Use AND when you want both terms included. For example, *performance AND lizards*. In this example, all papers that mention performance AND lizards in the title, abstract or keywords are returned. 
- Use OR when you want to include either term. For example, *performance AND lizards OR geckos*. This example should return more literature than the previous search as it will include, all papers that mention *performance AND lizards OR geckos* in the title, abstract or keywords are returned. You will note that geckos are lizards, but people who write about geckos might not mention that they are lizards in the title, abstract or keywords, so you may need to use these extra search terms in order to pick up all of the relevant literature on lizards. If your research requires you to cover all lizards, then you might need a lot of search terms in the OR string in order to pick up all of the literature that covers lizard performance.
- Use NOT when you specifically want to exclude a keyword. Following our example above, the search term *performance AND reptiles NOT snakes* should give you all performance papers on lizards, but you may want to exclude turtles and crocodilians as well if they have a lot of hits.

### Proximity operators and parentheses

You can also use the term NEAR to indicate that two terms need to occur within a specified number of words. For example, performance NEAR/5 reptile will capture items where these terms occur within 5 words of each other, but no greater. This might be useful to exclude some papers that mention performance and reptiles in the abstract, but are not about reptile performance.

Once you start putting together search strings you will need to use brackets to make sure that the correct terms are grouped. For example, performance AND (lizards OR geckos) will get what we described above, whereas (performance AND lizards) OR geckos will give you all papers on lizards and performance plus all papers on geckos (not what we wanted).

@@

### Wildcards

@@

### Combining searches

@@

### Citations

@@

#### What else are literature databases used for?

The Science Citation Index used to be published as a single volume that emerged each year and was available in the reference section of the library. By the time that I started at university as an undergraduate in the 1980s, it was several volumes large and my first task for my tutor (one Dr Dave J Thompson) was to look up all of his citations (thankfully not too many at that time). Within a couple of years, the database was digitised and available on CD, and eventually became accessible through the internet.

Citations provide an idea of how the literature is used. This can be used to assess not simply the productivity of a scientist (i.e. number of publications), but their usefulness by others (i.e. number of citations).

@@

## Administrators want metrics

@@

## Searching by scientist's name

Once you have found a scientist who works within your field and produces influential work that you are keen to read, then you may want to search for all of their other papers to determine whether there is more literature that you have missed. The databases make such searches easy, but you will need the (correct) spelling of your scientist's name, and also to be aware that they may have used different iterations of their name throughout their career.

### Overcoming the difficulties of common names

While Dave Thompson could always add his middle initial to specify his work, as time went by there were more and more scientists publishing as D J Thompson. This made it more difficult for administrators and scientits alike. This has begun to be a bigger global problem, such that the databases have begun to use profiles of particular scientists that they can 'own' and curate. Each database has its own way of doing this, and it can become rather a chore so that not many scientists have their own profiles, which makes it a problem particularly for those with common names.

The [ORCID](orcid.html) number has been errected to provide a common platform for [authors to curate](title.html), and now some journals and funding authorities won't allow you to submit or apply without one.

## Google Scholar, Web of Science or Scopus?

@@

I have been known to take the odd peak at my Google Scholar profile over the year, and see how the citations are accumulating. I rarely check on WoS or Scopus, because it's a bit of a faff getting signed in and doing the search. Plus it looks so much smaller when one is habituated to seeing those double digits in GS! However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among others that they give) as I've never really known what all that extra represents. Something grey and unseemly? Well, it turns out that it's all good stuff, and perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.

However, it is worth noting that of the three indices, Google Scholar seems to be the easiest to manipulate. Delgado López‐Cózar et al [-@delgadolopez-cozar2014google] show how it is possible to manipulate citations and therefore metrics of individual researchers by depositing false documents into institutional repositories (note that they did this as an experiment, and you may find that restrictions on doing this at your own institution are prohibitive or lax). The authors also note that a consistent problem with Google Scholar is the lack of transparency, and therefore verification, of how the bibliometrics are calculated. This will make a difference to what papers are found, as Google Scholar tends to report those with the greatest citations first on the first page of outputs. Hence once a paper is initially boosted by manipulation, it will be more likely to be found and increasingly cited.

@@

## What’s the difference between indices?

According to Martín-Martín et al [-@martin-martin2018google] The GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus. If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world. Google Scholar results reflect a more inclusive citation index [@martin-martin2018google]. While WoS and Scopus aren't exclusively English or journal publications, they are mostly. But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream. Is your GS score more than a third higher than your WoS or Scopus score? If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.
